The state of knowledge for marine invertebrate biodiversity in the continental US

### 2017-08-14 manuscript.Rmd
edit manuscript

Regional faunistic lists assembled and curated by taxonomic experts provide a comprehensive and detailed account of species that are known to occur in a given geographic area. We assessed how faunistic lists assembled over decades by specialists compare with records in public databases for the same geographic area.

<!-- need to separate better baises for which I have data and those for which I don't -->
Species that are included in the list but do not have records in databases could be (1) rare: they have never been collected or have been collected only a few times, and the corresponding records have not been digitized yet; (2) difficult to preserve: large, very small, or fragile species are deposited in museums less frequently; (3) are in the database under a different name: the species has been synonymized or re-identified since the publication of the list under a different name that may be included in the database; (4) at the limit of their range: the species were included in the list because they are known to occur within the boundaries of the geographic area, but they have never been collected within these geographic boundaries; (5) do not have geographic coordinates associated with them: some of these species can be in the database but without georeferenced data points and therefore were not returned in our queries.

Untangling the relative contribution of these factors to explain the discrepancies is difficult as it would require to know the criteria used by the experts who have assembled the lists. However, by querying iDigBio we found that for the `r format(not_really_in_database$idig_gom_n_spp_not_in_db, big.mark=",")` species that do not appear within the geographic range of the Gulf of Mexico but are on the list, `r format_output(not_really_in_database$idig_gom_p_spp_not_in_db*100)`% are completely absent from iDigBio across all areas, while for the `r format(not_really_in_database$idig_gom_n_spp_no_coords, big.mark=",")` species that appear in iDigBio, `r format_output(not_really_in_database$idig_gom_p_spp_no_coords*100)`%  do not have geographic coordinates associated with them, and for the ones that do have geographic coordinates, `r format_output(not_really_in_database$idig_gom_p_spp_within_usa*100)`% ($n=$  `r format(not_really_in_database$idig_gom_n_spp_within_usa, big.mark=",")`) fall within the US waters. These numbers indicate that the lack of geographic coordinates associated with the specimens in the database contribute at least significantly to the discrepancy, but that a large fraction of these species have no digital records of their occurrence in the Gulf of Mexico. In other words, additional digitization of collections (including capturing their geographic coordinates information) are needed. For instance, one can expect that the digitization coverage for the Gulf of Mexico will greatly increase once records from the Florida Wildlife Research Institute collections (St Petersburg, FL) become integrated into biodiversity information systems.

Species that are in the database but not in the faunistic list could be: (1) records not considered by the authors of the checklist: either because they were not accessible or known to the authors; (2) records erroneously excluded by the author of the checklist; (3) records that have been discovered since the list has been assembled; (4) records with incorrect identification: records identified as a species that does not occur in the area; (5) records with incorrect geographic coordinates: their associated geographic information in the database places them in the Gulf of Mexico or the Pacific Northwest, but they were actually collected elsewhere.

We investigated the proportion of species discovered since the publication of the faunistic lists using iDigBio and OBIS records. They represent `r sum(recently_collected$gom_in_dbs$min_year > 2001)` species for the Gulf of Mexico, and `r sum(recently_collected$pnw_in_dbs$min_year > 1996)` species for the Pacific Northwest. While errors in geographic coordinates cannot be completely ruled out, it is more likely that the main contributing factor was that the authors of these checklists were not aware of the existence of these species based on museum records, as they were difficult to discover before digitization which has occured after the compilation of the list in most cases. However, museum specimens are critical to validate the occurrence of species [@Rocha2014], especially for marine invertebrates for which species limits based on morphology have been known to be challenged by other lines of evidence including molecular data [@Knowlton2000].

@@

<!-- add sentence -->
<!-- separate idigbio and (and add) obis in the results -->
We combined `r format(nrow(idigbio_records), big.mark=",")` museum records for marine invertebrates collected within the EEZ of the continental US representing `r length(unique(tolower(idigbio_records[["phylum"]])))` phyla and `r format(length(unique(idigbio_records$worms_valid_name)), big.mark=",")` species from iDigBio. Most samples were collected on the continental shelf (Fig. \@ref(fig:map-diversity-sampling)A). For the East coast, species richness is the highest in South Florida but it is relatively uniform along the coastline. For the West coast, the Pacific Northwest, the San Francisco Bay Area, and Southern California have the highest richness (Fig. \@ref(fig:map-diversity-sampling)B).

```{r map-diversity-sampling, fig.cap='Spatial distribution of the number of samples (A) and species richness (B) for marine invertebrates based on records retrieved from iDigBio and OBIS within the limits of the Exclusive Economic Zone for the continental USA. (\\#fig:map-diversity-sampling)'}
knitr::include_graphics("figures/map_diversity_sampling.pdf")
```

Spatial patterns of species richness recovered are strongly confounded by sampling effort. Samples of <100 records per cell commonly have as many species as records (Fig. \@ref(fig:correlation-sampling)).

@@

Underscoring the inadequacy of the sampling effort, `r prop_singleton_species`% of the species are known from a single record in iDigBio. Another indication of this inadequacy is the relation between the number of collected specimen and the number of species recorded through time (Fig. \@ref(fig:idigbio-through-time)). Initial explorations in the late ninetieth century led to an initial exponential growth in the knowledge of the fauna, that tapered off, until the 1970s when intensive collecting efforts led to an increase in the recording of new taxa (Fig. \@ref(fig:idigbio-through-time)). Since then, the slowing down of collections has been accompanied with a slowing down in the discovery of new records: `r n_spp_comparison$n_spp_2017_2000` species have been documented for the first time since 2000 (`r format_output(n_spp_comparison$n_spp_2017_2000/18)` species.year^-1^), while `r n_spp_comparison$n_spp_1970_1960` have been documented between 1960 and 1970 (`r format_output(n_spp_comparison$n_spp_1970_1960/11)` species.year^-1^). The strong correlations between (1) the number of samples collected and the species richness (Fig. \@ref(fig:correlation-sampling)), and (2) the number of specimens collected per year and the number of species recorded for the first time (Fig. \@ref(fig:idigbio-through-time)C), suggest that any new sampling effort will lead to new observations and new species. <!-- separate issues of sampling is needed, and large-scale sampling is best way of doing it --> Coordinated large-scale biodiversity expeditions continue to be needed to improve our understanding of the diversity of species and their spatial distributions. The US marine EEZ is relatively well-known compared to many areas of the world ocean, where studies show that undescribed biodiversity dominates. For example, a recent survey in Western Australia reported that 94.6% of crustaceans and 72% of polychaetes taken at 135 sites were undescribed [@Poore2014].

@@

>>>>

<!-- Regional faunistic lists assembled and curated by taxonomic experts provide a comprehensive and detailed account of species that are known to occur in a given geographic area. We assessed how faunistic lists assembled over decades by specialists compare with records in public databases for the same geographic area. -->

Several factors may contribute for species to be included in the regional lists but to not have records in biodiversity databases. The first category of biases are linked to property of the species affecting the likelihood of having specimens preserved in museum collections. For instance, rare species may never have been collected or have only been collected a few times, and the corresponding records have not been digitized yet. Species that are difficult to preserve because they are large, very small, or fragile, are deposited in museums less frequently. The other category of biases is associated with data quality for these records. Some of these species may be in the databases but do not have geographic coordinates associated with them, and therefore would not be included in the results of our query.

Missing geographic information contribute to the discrepencies observed.  Using iDigBio we found that for the `r format(not_really_in_database$idig_gom_n_spp_not_in_db, big.mark=",")` species that do not appear within the geographic range of the Gulf of Mexico but are on the list, `r format_output(not_really_in_database$idig_gom_p_spp_not_in_db*100)`% are completely absent from iDigBio, while for the `r format(not_really_in_database$idig_gom_n_spp_no_coords, big.mark=",")` species that do appear in iDigBio, `r format_output(not_really_in_database$idig_gom_p_spp_no_coords*100)`%  do not have geographic coordinates associated with them, and for the ones that do have geographic coordinates, `r format_output(not_really_in_database$idig_gom_p_spp_within_usa*100)`% ($n=$  `r format(not_really_in_database$idig_gom_n_spp_within_usa, big.mark=",")`) fall within the US waters. These numbers indicate that missing geographic coordinates reduces the overlap between the list of species from the regional list and the databases. Additional digitization of collections, including the capture of their geographic coordinates information, are needed to improve biodiversity knowledge at regional scales. For instance, one can expect that the digitization coverage for the Gulf of Mexico will greatly increase once records from the Florida Wildlife Research Institute collections (St Petersburg, FL) become integrated into biodiversity information systems.

Two factors are most likely to contribute to explain why some species are in databases but were not considered by the authors of the checklists: (1) the species have been documented for the first time since the publication of the checklist, (2) the species records were not accessible or known to the authors.

<!-- include also proportion of species not in list -->
Recently recorded species contribute modestly to this discrepency, as `r sum(recently_collected$gom_in_dbs$min_year > 2001)` species (`r format_output( sum(recently_collected$gom_in_dbs$min_year > 2001)/n_distinct(total_gom_records$worms_valid_name) * 100)`% of the total species richness) have been collected for the first time since the publication of the list in the Gulf of Mexico, and `r sum(recently_collected$pnw_in_dbs$min_year > 1996)` species (`r format_output(sum(recently_collected$pnw_in_dbs$min_year > 1996)/n_distinct(total_pnw_records$worms_valid_name) * 100)`% of the total species richness) for the Pacific Northwest.

The lack to tools for the authors of these checklist to have access to museum records across institutions contributed most to explain why some species were not included in the checklists. These records were difficult to discover before digitization which, for the most part, has occured after the compilation of these checklists in most cases.

@@

<!-- add sentence -->
<!-- separate idigbio and (and add) obis in the results -->

We retrieved `r format(n_distinct(combined_species$worms_valid_name), big.mark = ",")` species (`r n_distinct(na.omit(combined_species$worms_phylum))` phyla) of marine invertebrates in the continental US EEZ by combining  `r format(nrow(idigbio_records), big.mark=",")` museum records from iDigBio (`r format(n_distinct(idigbio_records$worms_valid_name), big.mark = ",")` species, `r n_distinct(na.omit(idigbio_records$worms_phylum))` phyla), and `r format(nrow(obis_records), big.mark=",")` records from OBIS (`r format(n_distinct(obis_records$worms_valid_name), big.mark=",")` species,
`r n_distinct(na.omit(obis_records$worms_phylum))` phyla). Extant marine invertebrates occurring in the US represent less than 0.4% of the total iDigBio records, as it currently holds records for over 106 million specimens, and they represent about 1% of the records that occur in the United States (about 33 million records in iDigBio). Marine invertebrates that occur in the US represent approximately 4% of the 48 million occurrence records in OBIS.

<!-- need to add latitudinal gradient map -->
Most samples have been collected on the continental shelf (Fig. \@ref(fig:map-diversity-sampling)A). On the East coast, species richness is highest at low latitudes, while it increases towards high latitude on the West coast (Fig. \@ref(fig:map-diversity-sampling)B).

```{r map-diversity-sampling, fig.cap='Spatial distribution of the number of samples (A) and species richness (B) for marine invertebrates based on records retrieved from iDigBio and OBIS within the limits of the Exclusive Economic Zone for the continental USA. (\\#fig:map-diversity-sampling)'}
knitr::include_graphics("figures/combined_map_sampling.pdf")
```

Spatial patterns of species richness are strongly confounded by sampling effort. Samples of <100 records per cell commonly have as many species as records (Fig. \@ref(fig:correlation-sampling)).

@@

Underscoring the inadequacy of the sampling effort, `r prop_singleton_species`% of the species are known from a single record. Another indication of this inadequacy is the relation between the number of collected specimen and the number of species recorded through time (Fig. \@ref(fig:idigbio-through-time)). Initial explorations in the late ninetieth century led to an initial exponential growth in the knowledge of the fauna, that tapered off, until the 1970s when intensive collecting efforts led to an increase in the recording of new taxa (Fig. \@ref(fig:idigbio-through-time)). Since then, the slowing down of collections has been accompanied with a slowing down in the discovery of new records: `r n_spp_comparison$n_spp_2017_2000` species have been documented for the first time since 2000 (`r format_output(n_spp_comparison$n_spp_2017_2000/18)` species.year^-1^), while `r n_spp_comparison$n_spp_1970_1960` have been documented between 1960 and 1970 (`r format_output(n_spp_comparison$n_spp_1970_1960/11)` species.year^-1^). The strong correlations between (1) the number of samples collected and the species richness (Fig. \@ref(fig:correlation-sampling)), and (2) the number of specimens collected per year and the number of species recorded for the first time (Fig. \@ref(fig:idigbio-through-time)C), suggest that any new sampling effort will lead to new observations and new species. <!-- separate issues of sampling is needed, and large-scale sampling is best way of doing it --> Coordinated large-scale biodiversity expeditions continue to be needed to improve our understanding of the diversity of species and their spatial distributions. The US marine EEZ is relatively well-known compared to many areas of the world ocean, where studies show that undescribed biodiversity dominates. For example, a recent survey in Western Australia reported that 94.6% of crustaceans and 72% of polychaetes taken at 135 sites were undescribed [@Poore2014].

@@

 However, museum specimens are critical to validate the occurrence of species [@Rocha2014], especially for marine invertebrates for which species limits based on morphology have been known to be challenged by other lines of evidence including molecular data [@Knowlton2000].
